Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2019-12-11 article-information-2019.tex
ore edits, some results

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. While this manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems for high-stakes, human-centered, or regulated ML applications.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

\begin{table}[H]
\label{tab:sim_acc}
\caption{Accuracy metrics for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Model}		& \textbf{Accuracy}	& \textbf{AUC}	 & \textbf{Logloss}	& \textbf{RMSE}\\
\midrule
$g^\text{GBM}$		& 				& 			& 				& \\
$g^\text{MGBM}$		& 				& 			& 				& \\
$g^\text{ANN}$		& 				& 			& 				& \\
$g^\text{XNN}$		& 				& 			& 				& \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[H]
\label{tab:mort_acc}
\caption{Accuracy metrics for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the mortgage test data.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Model}		& \textbf{Accuracy}	& \textbf{AUC}	 & \textbf{Logloss}	& \textbf{RMSE}\\
\midrule
$g^\text{GBM}$		& 				& 			& 				& \\
$g^\text{MGBM}$		& 				& 			& 				& \\
$g^\text{ANN}$		& 				& 			& 				& \\
$g^\text{XNN}$		& 				& 			& 				& \\
\bottomrule
\end{tabular}
\end{table}

@@

\item \textbf{Model documentation}: models cards \cite{model_cards}, e.g. \texttt{\url{https://github.com/openai/gpt-2}}, \texttt{\url{https://modelcards.withgoogle.com/object-detection}}.

@@

\subsection{Model Transparency, Appeal, and Compliance}

Interpretable model architectures and post-hoc explanations play an important role in increasing transparency into model mechanisms and predictions. As seen in Sections \ref{sec:m_and_m} and \ref{sec:res}, interpretable models often enable users to enforce domain knowledge-based constraints on model behavior, to ensure that models obey reasonable expectations, and to gain data-derived insights into the modeled problem domain. Post-hoc explanations generally help summarize model mechanisms and decisions. Together they allow human learning from ML, certain types of regulatory compliance, and crucially, logical human appeal or override of automated model decisions \cite{art_and_sci}. Interpretable models and post-hoc explanations are likely good candidates for ML uses cases under the FCRA, ECOA, GDPR and other regulations that may require explanations of model decisions, and they are already used in the financial services industry today for model validation and other purposes.\footnote{See: \href{https://ww2.amstat.org/meetings/jsm/2019/onlineprogram/AbstractDetails.cfm?abstractid=303053}{Deep Insights into Explainability and Interpretability of Machine Learning Algorithms and Applications to Risk Management}.}\textsuperscript{,}\footnote{Unfortunately, many non-consistent explanation methods can result in drastically different global and local feature importance values across different models trained on the same data or even for refreshing a model with augmented training data \cite{molnar}. Consistency and accuracy guarantees are perhaps a factor in the growing momentum behind Shapley values as a candidate technique for generating consumer-specific adverse action notices for explaining and appealing automated ML-based decisions in highly-regulated settings such as credit lending \cite{bracke2019machine}.} Writ large, transparency in ML also facilitates additional responsible AI processes such as model debugging, model documentation, and logical appeal and override processes, some which may also be require by applicable regulations.\footnote{E.g.: \href{https://www.federalreserve.gov/supervisionreg/srletters/sr1107.htm}{US Federal Reserve Bank Supervision and Regulation (SR) Letter 11-7: Guidance on Model Risk Management}.} Among these, appeal may deserve the most attention. ML model are often wrong.\footnote{``All models are wrong, but some are useful.'' -- George Box, Statistician (1919 - 2013)} For high-stakes, human-centered, or regulated applications that are trusted with mission- or life-critical decisions, the ability to appeal or override inevitable wrong decisions is not only a frequent prerequisite for regulatory compliance, but also an important failsafe procedure for those affected by ML decisions.

@@

\subsection{Intersectionality of Interpretability, Fairness, and Security in ML}

The black-box nature of ML, the perpetuation or exacerbation of discrimination by ML, or the security vulnerabilities inherent in ML are each serious and difficult problems on their own. However, evidence is mounting that these harms can also manifest as complex intersectional challenges e.g. the \textit{fairwashing} or \textit{scaffolding} of biased models with ML explanations, the privacy harms of ML explanations, or the adversarial poisoning of ML models to become discriminatory \cite{fair_washing}, \cite{scaffolding}, \cite{shokri2019privacy}. \footnote{See: \href{https://www.telegraph.co.uk/technology/2016/03/24/microsofts-teen-girl-ai-turns-into-a-hitler-loving-sex-robot-wit/}{\textit{Microsoft deletes 'teen girl' AI after it became a Hitler-loving sex robot within 24 hours}}.}\textsuperscript{,}\footnote{While the focus of this paper is not ML security, proposed best-practices from that field do point to transparency of ML systems as a mitigating factor for some ML attacks and hacks \cite{papernot2018marauder}. High system complexity is sometimes considered a mitigating influence as well \cite{hoare19811980}. This is sometimes known as the \textit{transparency paradox} in data privacy and security, and it likely applies to ML security as well, especially in the context of interpretable ML models and post-hoc explanation.\textsuperscript{17}}\textsuperscript{,}\footnote{See: \textcolor{red}{Andrew Burt HBR article - upcoming.}} Again, Sections \ref{sec:m_and_m} and \ref{sec:res} make no claims that the opacity, discrimination, or security problems in ML have been solved, even treated as independent problems. Instead, this text aims to highlight these issues as both singular entities and intersectional phenomena. Practitioners should of course consider the discussed interpretable modeling, post-hoc explanation, and discrimination testing approaches as at least partial remedies to the black-box, and discrimination issues in ML. However, they should also consider that explanations can ease model stealing, data extraction, and membership inference attacks and that explanations can mask ML discrimination. Additionally, high-stakes, human-centered, or regulated ML systems should generally be built and tested with robustness to adversarial attacks as a primary design consideration, and specifically to prevent ML models from being poisoned or otherwise altered to become discriminatory. Of course, accuracy, discrimination, and security characteristics of a system can change over time too, and simply testing for these problems at training time, as presented in Sections \ref{sec:m_and_m} and \ref{sec:res}, is not adequate for high-stakes, human-centered, or regulated ML systems. Accuracy, discrimination, and security should be monitored in real-time and over time, as long as a model is deployed.

@@

This text puts forward results on simulated data to provide a rough validation of constrained ML models, post-hoc explanation techniques, and discrimination testing methods. These same modeling, explanation, and discrimination testing approaches are then applied to more realistic mortgage data to provide an example of a responsible machine learning work flow high-stakes, human-centered, or regulated ML applications. The discussed methodologies are solid steps toward interpretability, explanation, and minimal discrimination for ML decisions, which should ultimately enable increased fairness and logical appeal processes for ML decision subjects. Of course there is more to the responsible practice of ML than interpretable models, post-hoc explanation, and discrimination testing, even from a technology perspective, and Section \ref{sec:disc} also points out numerous additional references and open source Python software assets that are available to researchers and practitioners today to increase human trust and understanding in ML systems. While the messy, complex, and human problems of racism, sexism, privacy violations, and cyber crime can probably not be solved by technology alone, this work and many others illustrate numerous ways for ML practitioners to become part of the solution to these problems, instead of perpetuating and exacerbating them.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{Wen Phan for work in formalizing notation. \textcolor{red}{BLDS editor} for editing.} Andrew Burt for ideas around ML transparency, model debugging, and the transparency paradox.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer and investor of H2O.ai and a customer of BLDS, LLC.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this text: AI -- artificial intelligence, ANN -- artificial neural network, DI -- disparate impact, EBM or GA\textsuperscript{2}M -- explainable boosting machine, i.e. variants GAMs that consider two-way interactions and incorporate boosting into training, GAM -- generalized additive model, GBM -- gradient boosting machine, ICE -- individual conditional expectation, MGBM -- monotonic gradient boosting machine, ML -- machine learning, PD -- partial dependence, SGD -- stochastic gradient descent, SR -- Supervision and Regulation, US -- United States, XNN -- explainable neural network.}

>>>>

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. While this manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems for high-stakes, human-centered, or regulated ML applications.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

\begin{table}[H]
\label{tab:sim_acc}
\caption{Accuracy metrics for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Model}		& \textbf{Accuracy}	& \textbf{AUC}	 & \textbf{Logloss}	& \textbf{RMSE}\\
\midrule
$g^\text{GBM}$		& 0.775	& 0.857	& 0.474	& 0.394 \\
$g^\text{MGBM}$	& 0.763	& 0.846	& 0.498	& 0.405 \\
$g^\text{ANN}$		& 		& 			& 				& \\
$g^\text{XNN}$		& 		& 			& 				& \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[H]
\label{tab:mort_acc}
\caption{Accuracy metrics for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the mortgage test data.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Model}		& \textbf{Accuracy}	& \textbf{AUC}	 & \textbf{Logloss}	& \textbf{RMSE}\\
\midrule	
$g^\text{GBM}$		& 0.795	& 0.828	& 0.252	& 0.276 \\
$g^\text{MGBM}$	& 0.765	& 0.814	& 0.259	& 0.278 \\
$g^\text{ANN}$		& 		& 		& 		& \\
$g^\text{XNN}$		& 		& 0.868	& 0.233	& 0.263 \\
\bottomrule
\end{tabular}
\end{table}

@@

\item \textbf{Model documentation}: models cards \cite{model_cards}, e.g. \texttt{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}, \texttt{\url{https://modelcards.withgoogle.com/object-detection}}.

@@

\subsection{Interpretability, Explainability, Appeal, and Compliance}

Interpretable model architectures and post-hoc explanations play an important role in increasing transparency into model mechanisms and predictions. As seen in Sections \ref{sec:m_and_m} and \ref{sec:res}, interpretable models often enable users to enforce domain knowledge-based constraints on model behavior, to ensure that models obey reasonable expectations, and to gain data-derived insights into the modeled problem domain. Post-hoc explanations generally help describe and summarize mechanisms and decisions, potentially yielding an even clearer understanding of ML models. Together they can allow for human learning from ML, certain types of regulatory compliance, and crucially, human appeal or override of automated model decisions \cite{art_and_sci}. Interpretable models and post-hoc explanations are likely good candidates for ML uses cases under the FCRA, ECOA, GDPR and other regulations that may require explanations of model decisions, and they are already used in the financial services industry today for model validation and other purposes.\footnote{See: \href{https://ww2.amstat.org/meetings/jsm/2019/onlineprogram/AbstractDetails.cfm?abstractid=303053}{Deep Insights into Explainability and Interpretability of Machine Learning Algorithms and Applications to Risk Management}.}\textsuperscript{,}\footnote{Unfortunately, many non-consistent explanation methods can result in drastically different global and local feature importance values across different models trained on the same data or even for refreshing a model with augmented training data \cite{molnar}. Consistency and accuracy guarantees are perhaps a factor in the growing momentum behind Shapley values as a candidate technique for generating consumer-specific adverse action notices for explaining and appealing automated ML-based decisions in highly-regulated settings such as credit lending \cite{bracke2019machine}.} Writ large, transparency in ML also facilitates additional responsible AI processes such as model debugging, model documentation, and logical appeal and override processes, some which may also be required by applicable regulations.\footnote{E.g.: \href{https://www.federalreserve.gov/supervisionreg/srletters/sr1107.htm}{US Federal Reserve Bank Supervision and Regulation (SR) Letter 11-7: Guidance on Model Risk Management}.} Among these, appeal may deserve the most attention. ML models are often wrong.\footnote{``All models are wrong, but some are useful.'' -- George Box, Statistician (1919 - 2013)} For high-stakes, human-centered, or regulated applications that are trusted with mission- or life-critical decisions, the ability to appeal or override inevitable wrong decisions is not only a possible prerequisite for regulatory compliance, but also an important failsafe procedure for those affected by ML decisions.

@@

\subsection{Intersectionality of Interpretability, Explainability, Discrimination, and Security in ML}

The black-box nature of ML, the perpetuation or exacerbation of discrimination by ML, or the security vulnerabilities inherent in ML are each serious and difficult problems on their own. However, evidence is mounting that these harms can also manifest as complex intersectional challenges, e.g. the \textit{fairwashing} or \textit{scaffolding} of biased models with ML explanations, the privacy harms of ML explanations, or the adversarial poisoning of ML models to become discriminatory \cite{shokri2019privacy}, \cite{fair_washing}, \cite{scaffolding}.\footnote{See: \href{https://www.theguardian.com/technology/2016/mar/24/tay-microsofts-ai-chatbot-gets-a-crash-course-in-racism-from-twitter}{\textit{Tay, Microsoft's AI chatbot, gets a crash course in racism from Twitter}}.}\textsuperscript{,}\footnote{While the focus of this paper is not ML security, proposed best-practices from that field do point to transparency of ML systems as a mitigating factor for some ML attacks and hacks \cite{papernot2018marauder}. High system complexity is sometimes considered a mitigating influence as well \cite{hoare19811980}. This is sometimes known as the \textit{transparency paradox} in data privacy and security, and it likely applies to ML security as well, especially in the context of interpretable ML models and post-hoc explanation.\textsuperscript{17}}\textsuperscript{,}\footnote{See: \textcolor{red}{Andrew Burt HBR article - upcoming.}} Again, this text makes no claims that the opacity, discrimination, or security problems in ML have been solved, even treated as independent problems. Instead, this text aims to highlight these issues as both singular entities and non-static intersectional phenomena. Practitioners should of course consider the discussed interpretable modeling, post-hoc explanation, and discrimination testing approaches as at least partial remedies to the black-box and discrimination issues in ML. However, they should also consider that explanations can ease model stealing, data extraction, and membership inference attacks and that explanations can mask ML discrimination. Additionally, high-stakes, human-centered, or regulated ML systems should generally be built and tested with robustness to adversarial attacks as a primary design consideration, and specifically to prevent ML models from being poisoned or otherwise altered to become discriminatory. Accuracy, discrimination, and security characteristics of a system can change over time as well. Simply testing for these problems at training time, as presented in Sections \ref{sec:m_and_m} and \ref{sec:res}, is not adequate for high-stakes, human-centered, or regulated ML systems. Accuracy, discrimination, and security should be monitored in real-time and over time, as long as a model is deployed.

@@

This text puts forward results on simulated data to provide a rough validation of constrained ML models, post-hoc explanation techniques, and discrimination testing methods. These same modeling, explanation, and discrimination testing approaches are then applied to more realistic mortgage data to provide an example of a responsible ML work flow for high-stakes, human-centered, or regulated ML applications. The discussed methodologies are solid steps toward interpretability, explanation, and minimal discrimination for ML decisions, which should ultimately enable increased fairness and logical appeal processes for ML decision subjects. Of course there is more to the responsible practice of ML than interpretable models, post-hoc explanation, and discrimination testing, even from a technology perspective, and Section \ref{sec:disc} also points out numerous additional references and open source Python software assets that are available to researchers and practitioners today to increase human trust and understanding in ML systems. While the messy, complex, and human problems of racism, sexism, privacy violations, and cyber crime can probably not be solved by technology alone, this work (and many, many others) illustrate numerous ways for ML practitioners to become part of the solution to these problems, instead of perpetuating and exacerbating them.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{Wen Phan for work in formalizing notation. \textcolor{red}{BLDS editor} for editing. Andrew Burt for ideas around ML transparency, model debugging, and the transparency paradox.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer of, and investor in, H2O.ai and a customer of BLDS, LLC.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this text: 
AI -- artificial intelligence, 
AIR - adverse impact ratio,
ALE - accumulated local effect,
ANN -- artificial neural network, 
DI -- disparate impact, 
EBM or GA\textsuperscript{2}M -- explainable boosting machine, i.e. variants GAMs that consider two-way interactions and may incorporate boosting into training, 
ECOA - Equal Credit Opportunity Act,
EDA -- exploratory data analysis,
EU -- European Union,
FCRA -- Fair Credit Reporting Act,
GAM -- generalized additive model, 
GBM -- gradient boosting machine, 
GDPR - General Data Protection Regulation,
HMDA -- Home Mortgage Disclosure Act
ICE -- individual conditional expectation, 
MGBM -- monotonic gradient boosting machine, 
ML -- machine learning, 
PD -- partial dependence, 
SGD -- stochastic gradient descent, 
SHAP -- Shapley additive explanation,
SMD - standardized mean difference,
SR -- supervision and regulation, 
US -- United States, 
XNN -- explainable neural network.}
